Improved state-space models for inference about spatial and temporal variation in abundance from count data

### 2012-11-08 Dail_Mad_ext.tex
Map figure and simulation discussion

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

@@

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and, where appropriate, zero-inflated dynamics.  We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 tuning samples.
%Latent inclusion factors for temperature effects?

@@

@@

[simulation results discussion]

We have presented a test case for these models using data from
the BBS for two species. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
classical and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects and zero-inflated dynamics. Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have convergence problems,
particularly for the species (golden-winged warbler) with sparser data. It's
likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.
\hl{[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph
(currently not written), and golden-winged warbler ZIP results in ZIP
paragraph.]}

@@

>>>>

Theoretical ecology requires models of population dynamics for testing
hypotheses regarding spatial and temporal variation in
abundance. For example, much theoretical work has focused on
understanding the importance and existence of phenomenon such as
density-dependent population regulation, population cycling, and spatial synchrony
\citep{may:1975,royama:1977,turchin:1990,dennis_taper:1994,bjornstad_etal:1999},
and population models are required to evaluate associated hypotheses.
In applied contexts, population models are used for estimating extinction probabilities
\citep{schoener_spiller:1992,nadeem_lele:2011,hostetler_etal:2012} and for predicting the
effects of future environmental conditions or conservation actions on
population size \citep{jamieson_brooks:2004,hatfield_etal:2012}.
In order to address these questions, two complicating factors must
be confronted when fitting population models to data. 
First, deterministic models of population dynamics are virtually always inadequate due to process
variation, the inherent stochasticity in demographic parameters and environmental
conditions \citep{bjornstad_grenfell:2001,saether_engen:2002}.
Second, abundance---the natural state variable in studies
of population dynamics---can rarely be observed perfectly in field
studies because of observation error, such as imperfect
detection \citep{link_nichols:1994,kery_etal:2009}.

@@

We ran the top ranked models from the maximum likelihood
analyses in a Bayesian framework.  We added random observer 
effects and, where appropriate, zero-inflated dynamics.  We used
non-informative priors. We tested for lack of convergence using
5 Markov chains for each model \citep{gelman_rubin:1992}.
%(Gelman and Rubin 1992).
For each chain we sampled the MCMC for at least 40,000 iterations, after at
least 3,000 tuning samples.

We also used these models to obtain route and 
year specific abundance estimates.  Because of the memory requirements involved,
we obtained these estimates from a single chain of 5,000 iterations, after 35,000
tuning samples.  We present these estimates for ovenbirds only.
%Latent inclusion factors for temperature effects?

@@

We estimated ovenbird abundance by route and year (and averaged over routes) 
using the Bayesian model including random observer effects (Fig.~\ref{fig:oven_N}).  
Estimated average abundance oscillated over time with a slowly increasing trend.  Estimated ovenbird
abundance was highest in western and eastern Maryland in 1970 and 1980, but became
more evenly distributed by 2010.  Coefficient of variation (CV) for route and year specific
abundance estimates varied between 0.11 and 2.31 (mean = 0.46).  By comparison, CV
for the estimate of mean initial abundance ($\Lambda$) was 0.15.

@@

In our analysis of simulated data, we found that our models generally estimated parameters well.
Two exceptions were the estimates of instantaneous growth rate in exponential model when the 
true value of initial abundance was low and the estimates of equilibrium abundance and maximum
growth rate in the Ricker model when the true value of maximum growth rate was low.  
The failure of the first case was likely due to a large number of sites starting at abundance 0 or reaching it early in the simulation.  
%[Richard, maybe come up with a summary statistic for this based on the simulation N values?]
Since a site that hits abundance 0 in this simulation stayed at 0, this provided little information with
which to estimate growth rate.  The failure of the second case is also reasonable: with very
slow movement toward the equilibrium abundance comes little information for estimating
that abundance or the rate of growth.  In other cases, bias was low and coverage nominal.

We have presented a test case for these models using data from
the BBS for two species. We demonstrate that these models can be
used to model and estimate initial abundance, detection
probability, and density-independent and dependent functions for
population dynamics. We tested these models in both the
classical and Bayesian frameworks and found strengths and
weaknesses of each approach. The Bayesian analyses permit more
flexibility and allow the incorporation of random observer
effects and zero-inflated dynamics. Our results suggest ignoring
variation between observers can lead to underestimation of
initial abundance and equilibrium population sizes. On the other
hand, our Bayesian analyses did have convergence problems,
particularly for the species (golden-winged warbler) with sparser data. It's
likely that poor model fits are related to the low estimates of
detection probability, which in turn are probably caused by the
large 400 m radius used for BBS point counts. Again, we believe
that direct information on detectability (such as repeated
visits within a season) would greatly improve the fit of these
models in both frameworks.  We also were able to obtain year
and site-specific estimates of abundance (Fig.~\ref{fig:oven_N}),
although with generally lower precision than more global estimates.  These sort of estimates
may be more reliable when abundance and dynamics are modeled 
as functions of habitat covariates, which we did not do here.
\hl{[Alternatively, could put discussions of our results where
they fit in by subject area, rather than separately. For
instance, could discuss BBS population dynamics model selection
in population models paragraph, BBS Bayesian vs.
frequentist strengths and weakness in the Bayesian paragraph, 
and golden-winged warbler ZIP results in ZIP
paragraph.]}

@@

\begin{figure}
  \centering
  \includegraphics[width=6.6in]{figs/OVEN_N_by_route_year2}
\caption{Maps of BBS route and year specific estimated abundances for ovenbirds, for the years 1970,
1980, 1990, 2000, and 2010.  Lower right panel depicts estimated mean route abundance by year
(mean and 95\% credible interval).}
\label{fig:oven_N}
\end{figure}
